ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-27 pyked.tex
Last updates from me!

Instead, such data is typically found in under-documented comma-separated value (CSV)
files and Excel spreadsheets, or contained in PDF tables rather than as
supplementary material associated with a paper. In the worst case, data is only
available in figures and must be digitized by the use of software such as
WebPlotDigitizer~\autocite{WebPlotDigitizer}. These practices limit the wide use
of the valuable, extensive experimental data available in the literature.
Frenklach~\autocite{Frenklach:2007bm} further explained the benefits of
standardized, widely available data for combustion. In brief, much of
fundamental combustion\slash chemical kinetics research drives towards the
ultimate goal of developing predictive kinetic models---and ultimately this
depends on a \emph{community} infrastructure for data and methods.

@@

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make
it a strong standardization format, but it suffers from several flaws that
prevent wide adoption. First, the PrIMe standard does not require or support all
information needed to simulate an experiment, including a machine-readable
definition of ignition or a standard way to express detailed facility specific
effects important to properly simulating certain experiments. Second, the XML
format is intended to be a markup language rather than a data format, and
presents a barrier to creating and working with database files. Furthermore, the
focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for
scholarly products and InCHI or SMILES for species), limits usability outside
the PrIMe ecosystem. Finally, the closed and opaque nature of the PrIMe standard
and associated Data Warehouse limit contributions and community development of
the PrIMe standard and compatible data files and tools.

@@

In the \ck{} file, the experimental data is encoded in a sequence of mappings,
with the following information required in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping); The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured, one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point, one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the file
name of a \ck{} file or a Python dictionary containing the contents of a \ck{}
file as its argument. When the file or dictionary is loaded, it is validated to
have the correct format by the Python package Cerberus \autocite{cerberus}.

@@

The \pybox{DataPoint} class stores the information associated with a single data
point in the \ck{} file (i.e., a single element of the \yabox{datapoints}
sequence). Similar to the \pybox{ChemKED} class, information in the
\pybox{DataPoint} is stored as instance attributes:
%
\begin{itemize}
    \item \pybox{equivalence_ratio}: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the value
    \item \pybox{ignition_delay}, \pybox{temperature}, \pybox{pressure},
    \pybox{pressure_rise}: The value of these data are stored as instances of
    the \pybox{Quantity} class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \pybox{composition}: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yabox{mole-percent},
    \yabox{mole-fraction}, or \yabox{mass-fraction} cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yabox{mole-percent}.
    \item \pybox{composition_type}: A string indicating the type of composition
    information for the data point---one of \pybox{'mole-percent'},
    \pybox{'mole-fraction'}, or \pybox{'mass-fraction'}
    \item \pybox{volume_history}: If the \yabox{volume-history} of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \pybox{namedtuple}, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}.
    \item \pybox{compression_time}: If the \ck{} file encodes an RCM experiment,
    the length of the compression stroke may be included
\end{itemize}

@@

In this paper we presented the \ck{} data format for describing measurements
taken from fundamental combustion experiments, recognizing
that the community has a need for an open and standardized data serialization format.
\ck{} files are formatted using the YAML language and are plain-text, human- and
machine-readable, and easy to construct.

>>>>

Instead, such data is typically found in under-documented comma-separated value
(CSV) files and Excel spreadsheets, or contained in PDF tables rather than as
supplementary material associated with a paper. In the worst case, data is only
available in figures and must be digitized by the use of software such as
WebPlotDigitizer~\autocite{WebPlotDigitizer}. These practices limit the wide use
of the valuable, extensive experimental data available in the literature.
\textcite{Frenklach:2007bm} further explained the benefits of standardized,
widely available data for combustion. In brief, much of fundamental
combustion\slash chemical kinetics research drives towards the ultimate goal of
developing predictive kinetic models---and ultimately this depends on a
\emph{community} infrastructure for data and methods.

@@

PrIMe~\autocite{Frenklach:2007bm,You:2011hy} has a number of features that make
it a strong standardization format, but it suffers from several flaws that
prevent wide adoption. First, the PrIMe standard does not require or support all
the information needed to simulate an experiment, including a machine-readable
definition of ignition or a standard way to express detailed facility specific
effects important to properly simulating certain experiments. Second, the XML
format is intended to be a markup language rather than a data format, and
presents a barrier to creating and working with database files. Furthermore, the
focus on using internal identifiers for bibliographic references and
species\slash reactions, rather than standard identifiers (e.g., DOIs for
scholarly products and InCHI or SMILES for species), limits usability outside
the PrIMe ecosystem. Finally, the closed and opaque nature of the PrIMe standard
and associated Data Warehouse limit contributions and community development of
the PrIMe standard and compatible data files and tools.

@@

The second section of the file encodes the experimental data. In the \ck{} file,
this is done in a sequence of mappings, with the following information required
in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping); The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured, one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point, one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the file name
of a \ck{} file or a Python dictionary containing the contents of a \ck{} file
as its argument. When the file or dictionary is loaded, it is validated to have
the correct format by the Python package Cerberus \autocite{cerberus}. The
schema used for validation of the \ck{} files is available publicly in the
source code repository for
PyKED\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/chemked_schema.yaml}}.

@@

The \pybox{DataPoint} class stores the information associated with a single data
point in the \ck{} file (i.e., a single element of the \yabox{datapoints}
sequence). Similar to the \pybox{ChemKED} class, information in the
\pybox{DataPoint} is stored as instance attributes:
%
\begin{itemize}
    \item \pybox{equivalence_ratio}: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the
    value.
    \item \pybox{ignition_delay}, \pybox{temperature}, \pybox{pressure},
    \pybox{pressure_rise}: The value of these data are stored as instances of
    the \pybox{Quantity} class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \pybox{composition}: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yabox{mole-percent},
    \yabox{mole-fraction}, or \yabox{mass-fraction} cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yabox{mole-percent}.
    \item \pybox{composition_type}: A string indicating the type of composition
    information for the data point---one of \pybox{'mole-percent'},
    \pybox{'mole-fraction'}, or \pybox{'mass-fraction'}
    \item \pybox{volume_history}: If the \yabox{volume-history} of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \pybox{namedtuple}, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}
    \item \pybox{compression_time}: If the \ck{} file encodes an RCM experiment,
    the length of the compression stroke may be included
    \item \pybox{ignition_type}: A dictionary specifying the method of the
    measurement of ignition delay
\end{itemize}

@@

In this paper, we presented the \ck{} data format for describing measurements
taken from fundamental combustion experiments, recognizing that the community
has a need for an open and standardized data serialization format. \ck{} files
are formatted using the YAML language and are plain-text, human- and
machine-readable, and easy to construct.
